---
title: "Data Sources and Ingestion"
---

## Overview

This chapter fetches raw data from two public sources, loads it into DuckDB,
and verifies schema and coverage before analysis.

## Fetching Ookla Data via AWS CLI

The Ookla Open Dataset is hosted in a public S3 bucket — no AWS account needed.
Run once from the project root:

```bash
bash data-raw/fetch_ookla.sh
```

This downloads one Parquet file per quarter (~200 MB each) for 2022–2024:

```
data/ookla/
  year=2022/quarter=1/2022-01-01_performance_mobile_tiles.parquet
  year=2022/quarter=2/...
  ...
  year=2024/quarter=4/...
```

**Why AWS CLI here instead of R?**
`aws s3 cp --recursive` handles retries, partial downloads, and progress
reporting better than a bare `download.file()`. It also makes the data
provenance explicit — you can see exactly what bucket and path the files
came from.

## Loading into DuckDB with arrow

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(arrow)
library(dplyr)
library(sf)

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb")
```

```{r}
#| label: load-ookla
#| message: false
parquet_dir <- "../data/ookla"

if ("ookla_mobile" %in% dbListTables(con)) {
  message("Table already loaded — skipping.")
} else {
  ds <- open_dataset(parquet_dir, partitioning = c("year", "quarter"))
  ookla_df <- ds |>
    select(quadkey, avg_d_kbps, avg_u_kbps, avg_lat_ms,
           tests, devices, year, quarter) |>
    collect()
  dbWriteTable(con, "ookla_mobile", ookla_df)
  message("Loaded ", nrow(ookla_df), " rows into ookla_mobile")
}
```

## Schema and Coverage

```{sql connection=con}
DESCRIBE ookla_mobile;
```

```{sql connection=con}
SELECT
    year,
    quarter,
    COUNT(*)                          AS tiles,
    ROUND(AVG(avg_d_kbps) / 1000, 2) AS avg_down_mbps,
    ROUND(AVG(avg_u_kbps) / 1000, 2) AS avg_up_mbps,
    ROUND(AVG(avg_lat_ms), 1)         AS avg_latency_ms
FROM ookla_mobile
GROUP BY year, quarter
ORDER BY year, quarter;
```

## Data Dictionary

| Column | Type | Description |
|---|---|---|
| `quadkey` | VARCHAR | Bing Maps tile ID at zoom level 16 (~600m tiles) |
| `avg_d_kbps` | INTEGER | Mean download speed across all tests in tile (kbps) |
| `avg_u_kbps` | INTEGER | Mean upload speed (kbps) |
| `avg_lat_ms` | INTEGER | Mean latency (ms) |
| `tests` | INTEGER | Number of speed tests in tile |
| `devices` | INTEGER | Unique devices tested |
| `year` / `quarter` | INTEGER | Partition columns added at ingest |

## About Quadkeys

Ookla tiles use the [Bing Maps quadkey](https://learn.microsoft.com/en-us/bingmaps/articles/bing-maps-tile-system)
system at zoom level 16. Each tile covers roughly 600m × 600m at mid-latitudes.
The `sf` package can convert these to geometries for spatial joins.
