---
title: "Data Preparation"
---

## Overview

Raw Ookla tiles cover the entire world. This chapter filters to the US,
converts speeds to Mbps, derives time variables, and builds a clean
analysis table in DuckDB.

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(lubridate)
library(sf)

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb")
```

## What is a Quadkey?

Each row in the raw data is a geographic tile identified by a quadkey string —
a [Bing Maps quadkey](https://learn.microsoft.com/en-us/bingmaps/articles/bing-maps-tile-system)
at zoom level 16 (~600m tiles). Each of the 16 characters is a base-4 digit
(0–3) that encodes one bit of the tile's X coordinate and one bit of its Y
coordinate. Decoding all 16 digits gives `tile_x` and `tile_y`, which convert
to a longitude/latitude via the standard Web Mercator tile formula.

## Decoding Quadkeys to Coordinates and Filtering to the US

We decode `quadkey` to `tile_x`/`tile_y` (bit 0 of each digit → X, bit 1 → Y),
convert to the tile center `lon`/`lat`, and keep only tiles inside CONUS,
Alaska, or Hawaii bounding boxes. (This excludes Puerto Rico and other
territories — a known limitation, noted in the Discussion.)

```{r}
#| label: us-filter
dbExecute(con, "
  CREATE OR REPLACE TABLE ookla_us AS
  WITH coords AS (
      SELECT
          *,
          list_dot_product(
              list_transform(str_split(quadkey, ''), x -> CAST(x AS INTEGER) & 1),
              [32768,16384,8192,4096,2048,1024,512,256,128,64,32,16,8,4,2,1]
          ) AS tile_x,
          list_dot_product(
              list_transform(str_split(quadkey, ''), x -> (CAST(x AS INTEGER) >> 1) & 1),
              [32768,16384,8192,4096,2048,1024,512,256,128,64,32,16,8,4,2,1]
          ) AS tile_y
      FROM ookla_mobile
  ),
  lonlat AS (
      SELECT
          *,
          360.0 * ((tile_x * 256 + 128) / 16777216.0 - 0.5) AS lon,
          90 - 360 * ATAN(EXP(-(0.5 - (tile_y * 256 + 128) / 16777216.0) * 2 * PI())) / PI() AS lat
      FROM coords
  )
  SELECT
      quadkey,
      lon,
      lat,
      avg_d_kbps / 1000.0                     AS avg_d_mbps,
      avg_u_kbps / 1000.0                     AS avg_u_mbps,
      avg_lat_ms,
      tests,
      devices,
      year,
      quarter,
      MAKE_DATE(year, (quarter - 1) * 3 + 1, 1) AS period_start
  FROM lonlat
  WHERE avg_d_kbps > 0
    AND tests >= 3
    AND (
          (lon BETWEEN -125    AND -66.5  AND lat BETWEEN 24   AND 49.5)  -- CONUS
       OR (lon BETWEEN -170    AND -129   AND lat BETWEEN 51   AND 72)    -- Alaska
       OR (lon BETWEEN -160.5  AND -154.5 AND lat BETWEEN 18.5 AND 22.5)  -- Hawaii
        )
")
cat("Rows in ookla_us:", dbGetQuery(con, "SELECT COUNT(*) AS n FROM ookla_us")$n)
```

## Speed Distribution

```{r}
#| label: speed-summary
dbGetQuery(con, "
  SELECT
      year,
      quarter,
      COUNT(*)                    AS tiles,
      ROUND(MIN(avg_d_mbps),  1)  AS min_dl,
      ROUND(AVG(avg_d_mbps),  1)  AS mean_dl,
      ROUND(MEDIAN(avg_d_mbps),1) AS median_dl,
      ROUND(MAX(avg_d_mbps),  1)  AS max_dl
  FROM ookla_us
  GROUP BY year, quarter
  ORDER BY year, quarter
")
```

## Removing Outliers

Speed tests above 2,000 Mbps on mobile are implausible (likely fixed broadband
mislabeled). We cap at the 99th percentile per period.

```{r}
#| label: outlier-removal
p99 <- dbGetQuery(con, "
  SELECT PERCENTILE_CONT(0.99) WITHIN GROUP (ORDER BY avg_d_mbps) AS p99
  FROM ookla_us
")$p99

dbExecute(con, glue::glue("
  CREATE OR REPLACE TABLE ookla_clean AS
  SELECT * FROM ookla_us
  WHERE avg_d_mbps <= {round(p99, 1)}
"))

cat("Rows after outlier removal:",
    dbGetQuery(con, "SELECT COUNT(*) FROM ookla_clean")[[1]])
```

## Period Labels

```{r}
#| label: period-labels
dbGetQuery(con, "
  SELECT DISTINCT year, quarter, period_start
  FROM ookla_clean
  ORDER BY year, quarter
")
```
