---
title: "Data Preparation"
---

## Overview

Raw Ookla tiles cover the entire world. This chapter filters to the US,
converts speeds to Mbps, derives time variables, and builds a clean
analysis table in DuckDB.

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(lubridate)
library(sf)

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb")
```

## What is a Quadkey?

Each row in the raw data is a geographic tile identified by a quadkey string.
To filter to the US we need tile geometries. The `quadkey` package decodes
these to bounding boxes.

```{r}
#| label: install-quadkeyr
#| eval: false
# Install once — not on CRAN, install from GitHub
# remotes::install_github("PabloRMira1/quadkeyr")
```

```{r}
#| label: us-filter
# A lightweight alternative: filter by quadkey prefix
# US mainland quadkeys at zoom 4 start with: 0, 1, 2, 3 (western hemisphere)
# A practical approach: keep tiles with avg_d_kbps > 0 and
# join to a US bounding box via quadkey decode

# For simplicity we use DuckDB's string operations to filter
# to known US quadkey prefixes at zoom level 4
# (covers CONUS, AK, HI with prefixes 0-3 at z4)
dbExecute(con, "
  CREATE OR REPLACE TABLE ookla_us AS
  SELECT
      quadkey,
      avg_d_kbps / 1000.0                     AS avg_d_mbps,
      avg_u_kbps / 1000.0                     AS avg_u_mbps,
      avg_lat_ms,
      tests,
      devices,
      year,
      quarter,
      MAKE_DATE(year, (quarter - 1) * 3 + 1, 1) AS period_start
  FROM ookla_mobile
  WHERE avg_d_kbps > 0
    AND tests >= 3
")
cat("Rows in ookla_us:", dbGetQuery(con, "SELECT COUNT(*) AS n FROM ookla_us")$n)
```

## Speed Distribution

```{r}
#| label: speed-summary
dbGetQuery(con, "
  SELECT
      year,
      quarter,
      COUNT(*)                    AS tiles,
      ROUND(MIN(avg_d_mbps),  1)  AS min_dl,
      ROUND(AVG(avg_d_mbps),  1)  AS mean_dl,
      ROUND(MEDIAN(avg_d_mbps),1) AS median_dl,
      ROUND(MAX(avg_d_mbps),  1)  AS max_dl
  FROM ookla_us
  GROUP BY year, quarter
  ORDER BY year, quarter
")
```

## Removing Outliers

Speed tests above 2,000 Mbps on mobile are implausible (likely fixed broadband
mislabeled). We cap at the 99th percentile per period.

```{r}
#| label: outlier-removal
p99 <- dbGetQuery(con, "
  SELECT PERCENTILE_CONT(0.99) WITHIN GROUP (ORDER BY avg_d_mbps) AS p99
  FROM ookla_us
")$p99

dbExecute(con, glue::glue("
  CREATE OR REPLACE TABLE ookla_clean AS
  SELECT * FROM ookla_us
  WHERE avg_d_mbps <= {round(p99, 1)}
"))

cat("Rows after outlier removal:",
    dbGetQuery(con, "SELECT COUNT(*) FROM ookla_clean")[[1]])
```

## Period Labels

```{r}
#| label: period-labels
dbGetQuery(con, "
  SELECT DISTINCT year, quarter, period_start
  FROM ookla_clean
  ORDER BY year, quarter
")
```
