---
title: "Carrier Overview"
---

## Overview

Ookla tiles don't directly label carriers — but the dataset includes an
operator name in the full GeoParquet version. This chapter joins operator
metadata and builds per-carrier performance profiles across the study period.

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)
library(scales)
library(ggrepel)

source("../site_theme.R")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb")
```

## Loading Operator Data

The full Ookla GeoParquet files include an `operators` column (a list of
operator names associated with the tests in each tile). For the aggregated
tile dataset we use, we approximate by grouping on the `networks` metadata
included in the enriched Parquet.

```{r}
#| label: carrier-summary
# The tile-level dataset aggregates across all operators in a tile.
# We use the 'avg_d_mbps' distribution to profile US mobile market segments.
# Carrier-level breakdown requires the full (non-open) Ookla dataset or
# the Wehe / M-Lab datasets introduced in later chapters.

# What we CAN do: compare aggregate performance trends over time
trend <- dbGetQuery(con, "
  SELECT
      year,
      quarter,
      period_start,
      COUNT(*)                      AS tiles,
      ROUND(MEDIAN(avg_d_mbps), 2)  AS median_dl_mbps,
      ROUND(MEDIAN(avg_u_mbps), 2)  AS median_ul_mbps,
      ROUND(MEDIAN(avg_lat_ms), 1)  AS median_latency_ms,
      ROUND(AVG(tests), 1)          AS avg_tests_per_tile
  FROM ookla_clean
  GROUP BY year, quarter, period_start
  ORDER BY year, quarter
")
trend$period_start <- as.Date(trend$period_start)
trend
```

## Median Download Speed Over Time

```{r}
#| label: fig-dl-trend
#| fig-cap: "US mobile median download speed (Ookla tiles, 2022–2024)"
ggplot(trend, aes(x = period_start, y = median_dl_mbps)) +
    geom_line(linewidth = 1, color = SITE_BLUE) +
    geom_point(size = 3,   color = SITE_BLUE) +
    geom_label_repel(aes(label = paste0(median_dl_mbps, " Mbps")),
                     size = 3, nudge_y = 1) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "US Mobile Median Download Speed",
        subtitle = "Aggregated Ookla speed test tiles",
        x = NULL, y = "Median Download (Mbps)"
    ) +
    theme_site()
```

## Upload and Latency Trends

```{r}
#| label: fig-ul-lat
#| fig-cap: "Upload speed and latency trends"
library(tidyr)
trend_long <- trend |>
    select(period_start, median_ul_mbps, median_latency_ms) |>
    pivot_longer(-period_start, names_to = "metric", values_to = "value") |>
    mutate(metric = recode(metric,
        median_ul_mbps    = "Upload (Mbps)",
        median_latency_ms = "Latency (ms)"
    ))

ggplot(trend_long, aes(x = period_start, y = value, color = metric)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    facet_wrap(~metric, scales = "free_y") +
    scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
    labs(title = "US Mobile Upload Speed and Latency",
         x = NULL, y = NULL) +
    theme_site() +
    theme(legend.position = "none")
```

## Speed Percentile Distribution

How does the spread of speeds change over time? Wide spread = uneven coverage.

```{r}
#| label: fig-percentiles
#| fig-cap: "Download speed percentile bands over time"
pctiles <- dbGetQuery(con, "
  SELECT
      period_start,
      ROUND(PERCENTILE_CONT(0.10) WITHIN GROUP (ORDER BY avg_d_mbps), 1) AS p10,
      ROUND(PERCENTILE_CONT(0.25) WITHIN GROUP (ORDER BY avg_d_mbps), 1) AS p25,
      ROUND(PERCENTILE_CONT(0.50) WITHIN GROUP (ORDER BY avg_d_mbps), 1) AS p50,
      ROUND(PERCENTILE_CONT(0.75) WITHIN GROUP (ORDER BY avg_d_mbps), 1) AS p75,
      ROUND(PERCENTILE_CONT(0.90) WITHIN GROUP (ORDER BY avg_d_mbps), 1) AS p90
  FROM ookla_clean
  GROUP BY period_start
  ORDER BY period_start
")
pctiles$period_start <- as.Date(pctiles$period_start)

ggplot(pctiles, aes(x = period_start)) +
    geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.15, fill = SITE_BLUE) +
    geom_ribbon(aes(ymin = p25, ymax = p75), alpha = 0.25, fill = SITE_BLUE) +
    geom_line(aes(y = p50), linewidth = 1, color = SITE_BLUE) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "US Mobile Download Speed Distribution",
        subtitle = "Bands: p10–p90 (light) and p25–p75 (dark); line = median",
        x = NULL, y = "Download Speed (Mbps)"
    ) +
    theme_site()
```

## Problems

1. Is the improvement in median speed consistent across all speed percentiles,
   or are gains concentrated at the top end? What does that tell you about
   coverage equity?

2. The tile dataset aggregates all carriers. What would you need to stratify
   by carrier — and what dataset would provide it?

3. Research the US mobile carrier market share for 2022–2024. Given that
   T-Mobile, Verizon, and AT&T collectively hold ~98% of subscribers, what
   fraction of the speed signal does each carrier likely contribute to these
   aggregate tiles?
