---
title: "Throughput Trends"
---

## Overview

This chapter models how throughput varies by time, geography, and
test volume — building intuition for what "normal" mobile performance
looks like before we search for throttling signals in Chapter 5.

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb")

df <- dbGetQuery(con, "SELECT * FROM ookla_clean")
df$period_start <- as.Date(df$period_start)
```

## Quarter-over-Quarter Change

```{r}
#| label: qoq-change
qoq <- dbGetQuery(con, "
  WITH medians AS (
      SELECT
          year, quarter, period_start,
          MEDIAN(avg_d_mbps) AS med_dl
      FROM ookla_clean
      GROUP BY year, quarter, period_start
  )
  SELECT
      year, quarter, period_start,
      med_dl,
      ROUND(med_dl - LAG(med_dl) OVER (ORDER BY period_start), 2) AS qoq_change,
      ROUND((med_dl - LAG(med_dl) OVER (ORDER BY period_start))
            / LAG(med_dl) OVER (ORDER BY period_start) * 100, 1)  AS qoq_pct
  FROM medians
  ORDER BY period_start
")
qoq$period_start <- as.Date(qoq$period_start)
qoq
```

## Speed vs. Test Density

More tests per tile should correlate with denser urban coverage — but are
denser tiles faster or slower?

```{r}
#| label: fig-tests-vs-speed
#| fig-cap: "Download speed vs. test count per tile (2024 Q4)"
df |>
    filter(year == 2024, quarter == 4, tests <= 500) |>
    ggplot(aes(x = tests, y = avg_d_mbps)) +
    geom_hex(bins = 60) +
    geom_smooth(method = "loess", color = "red", se = FALSE, linewidth = 0.8) +
    scale_fill_viridis_c(labels = comma) +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Download Speed vs. Test Count per Tile",
        subtitle = "2024 Q4 — hex bins, LOESS smoother",
        x = "Tests per Tile", y = "Avg Download Speed (Mbps)",
        fill = "Tiles"
    ) +
    theme_minimal()
```

## Speed Distribution by Year

```{r}
#| label: fig-ridge-dist
#| fig-cap: "Distribution of download speeds by year"
df |>
    mutate(year = factor(year)) |>
    ggplot(aes(x = avg_d_mbps, fill = year, color = year)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_x_continuous(limits = c(0, 300),
                       labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Mobile Download Speed Distribution by Year",
        x = "Download Speed (Mbps)", y = "Density",
        fill = "Year", color = "Year"
    ) +
    theme_minimal()
```

## High vs. Low Speed Tiles

Classify tiles into speed tiers and track their share over time:

```{r}
#| label: fig-speed-tiers
#| fig-cap: "Share of tiles by speed tier over time"
tiers <- df |>
    mutate(tier = case_when(
        avg_d_mbps <  10  ~ "< 10 Mbps (poor)",
        avg_d_mbps <  25  ~ "10–25 Mbps (basic)",
        avg_d_mbps < 100  ~ "25–100 Mbps (good)",
        TRUE              ~ "> 100 Mbps (excellent)"
    )) |>
    group_by(period_start, tier) |>
    summarise(n = n(), .groups = "drop") |>
    group_by(period_start) |>
    mutate(share = n / sum(n))

tier_order <- c("< 10 Mbps (poor)", "10–25 Mbps (basic)",
                "25–100 Mbps (good)", "> 100 Mbps (excellent)")
tiers$tier <- factor(tiers$tier, levels = tier_order)

ggplot(tiers, aes(x = period_start, y = share, fill = tier)) +
    geom_area(position = "stack", alpha = 0.85) +
    scale_y_continuous(labels = percent) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    scale_fill_brewer(palette = "RdYlGn") +
    labs(
        title = "US Mobile Speed Tier Distribution Over Time",
        x = NULL, y = "Share of Tiles", fill = "Tier"
    ) +
    theme_minimal()
```

## Problems

1. The FCC defines "broadband" as 25 Mbps down / 3 Mbps up (recently proposed
   to raise to 100/20). What share of tiles meet each threshold in each quarter?

2. Fit a simple linear regression of `avg_d_mbps ~ period_start`. What is the
   estimated quarterly improvement in Mbps? Is this statistically significant?

3. Compute the ratio `avg_d_mbps / avg_u_mbps` (download-to-upload ratio) per
   tile. How does this ratio vary by speed tier? What does an unusually high
   ratio suggest about the network?
