---
title: "Throughput Trends"
---

## Overview

This chapter models how throughput varies by time, geography, and
test volume — building intuition for what "normal" mobile performance
looks like before we search for throttling signals in Chapter 5.

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)

source("../site_theme.R")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb", read_only = TRUE)
```

## Quarter-over-Quarter Change

```{r}
#| label: qoq-change
qoq <- dbGetQuery(con, "
  WITH medians AS (
      SELECT
          year, quarter, period_start,
          MEDIAN(avg_d_mbps) AS med_dl
      FROM ookla_clean
      GROUP BY year, quarter, period_start
  )
  SELECT
      year, quarter, period_start,
      med_dl,
      ROUND(med_dl - LAG(med_dl) OVER (ORDER BY period_start), 2) AS qoq_change,
      ROUND((med_dl - LAG(med_dl) OVER (ORDER BY period_start))
            / LAG(med_dl) OVER (ORDER BY period_start) * 100, 1)  AS qoq_pct
  FROM medians
  ORDER BY period_start
")
qoq$period_start <- as.Date(qoq$period_start)
qoq
```

## Speed vs. Test Density

More tests per tile should correlate with denser urban coverage — but are
denser tiles faster or slower?

```{r}
#| label: fig-tests-vs-speed
#| fig-cap: "Download speed vs. test count per tile (2024 Q4)"
# Pre-bin both axes in SQL: 10-test buckets vs 5 Mbps buckets
df_q4 <- dbGetQuery(con, "
  SELECT
      FLOOR(tests / 10) * 10        AS test_bin,
      FLOOR(avg_d_mbps / 5) * 5     AS speed_bin,
      COUNT(*)                       AS n,
      AVG(avg_d_mbps)                AS avg_speed
  FROM ookla_clean
  WHERE year = 2024 AND quarter = 4 AND tests <= 500
  GROUP BY test_bin, speed_bin
  ORDER BY test_bin, speed_bin
")

# Smooth trend: mean speed per test bucket
trend_q4 <- df_q4 |>
    group_by(test_bin) |>
    summarise(avg_speed = weighted.mean(avg_speed, n), .groups = "drop")

ggplot(df_q4, aes(x = test_bin, y = speed_bin)) +
    geom_tile(aes(fill = n)) +
    geom_line(data = trend_q4, aes(x = test_bin, y = avg_speed),
              color = SITE_CORAL, linewidth = 1.2, inherit.aes = FALSE) +
    scale_fill_gradientn(colors = c("#f0ede8", "#b8d4f0", "#185FA5"),
                         labels = comma) +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Download Speed vs. Test Count per Tile",
        subtitle = "2024 Q4 — binned tile counts; red line = mean speed per bucket",
        x = "Tests per Tile", y = "Avg Download Speed (Mbps)",
        fill = "Tiles"
    ) +
    theme_site()
```

## Speed Distribution by Year

```{r}
#| label: fig-ridge-dist
#| fig-cap: "Download speed percentile spread by year (p10 / p25 / median / p75 / p90)"
dist_stats <- dbGetQuery(con, "
  SELECT
      year,
      PERCENTILE_CONT(0.10) WITHIN GROUP (ORDER BY avg_d_mbps) AS p10,
      PERCENTILE_CONT(0.25) WITHIN GROUP (ORDER BY avg_d_mbps) AS p25,
      PERCENTILE_CONT(0.50) WITHIN GROUP (ORDER BY avg_d_mbps) AS p50,
      PERCENTILE_CONT(0.75) WITHIN GROUP (ORDER BY avg_d_mbps) AS p75,
      PERCENTILE_CONT(0.90) WITHIN GROUP (ORDER BY avg_d_mbps) AS p90
  FROM ookla_clean
  GROUP BY year
  ORDER BY year
")
dist_stats$year <- factor(dist_stats$year)

ggplot(dist_stats, aes(x = year)) +
    geom_boxplot(
        aes(ymin = p10, lower = p25, middle = p50, upper = p75, ymax = p90),
        stat = "identity", fill = SITE_BLUE, color = "#0d3d6b",
        alpha = 0.45, width = 0.45
    ) +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Mobile Download Speed Distribution by Year",
        subtitle = "Box: p25–p75; whiskers: p10–p90; line: median",
        x = "Year", y = "Download Speed (Mbps)"
    ) +
    theme_site()
```

## High vs. Low Speed Tiles

Classify tiles into speed tiers and track their share over time:

```{r}
#| label: fig-speed-tiers
#| fig-cap: "Share of tiles by speed tier over time"
tiers <- dbGetQuery(con, "
  SELECT
      period_start,
      CASE
          WHEN avg_d_mbps <  10  THEN '< 10 Mbps (poor)'
          WHEN avg_d_mbps <  25  THEN '10–25 Mbps (basic)'
          WHEN avg_d_mbps < 100  THEN '25–100 Mbps (good)'
          ELSE '> 100 Mbps (excellent)'
      END AS tier,
      COUNT(*) AS n
  FROM ookla_clean
  GROUP BY period_start, tier
  ORDER BY period_start
")
tiers$period_start <- as.Date(tiers$period_start)
tiers <- tiers |>
    group_by(period_start) |>
    mutate(share = n / sum(n)) |>
    ungroup()

tier_order <- c("< 10 Mbps (poor)", "10–25 Mbps (basic)",
                "25–100 Mbps (good)", "> 100 Mbps (excellent)")
tiers$tier <- factor(tiers$tier, levels = tier_order)

ggplot(tiers, aes(x = period_start, y = share, fill = tier)) +
    geom_area(position = "stack", alpha = 0.85) +
    scale_y_continuous(labels = percent) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    scale_fill_manual(values = SPEED_TIER_COLORS) +
    labs(
        title = "US Mobile Speed Tier Distribution Over Time",
        x = NULL, y = "Share of Tiles", fill = "Tier"
    ) +
    theme_site()
```
