---
title: "Throttling Detection"
---

## Overview

Carriers typically advertise "unlimited" data but state in fine print that
speeds may be reduced after a usage threshold (commonly 50–100 GB/month).
This chapter introduces the statistical methods for detecting such breakpoints
and applies them to throughput time series.

**Important caveat:** The Ookla tile dataset aggregates all tests in a tile
across all users — it cannot directly observe a single user's data consumption.
What we *can* detect:

- Structural breaks in aggregate throughput (e.g. after carrier policy changes)
- Tiles where throughput is persistently below expected given network density
- Anomalous quarter-over-quarter drops that don't match national trends

True per-user throttle detection requires the Wehe or M-Lab datasets
(introduced at the end of this chapter).

## Setup

```{r}
#| label: setup
#| include: false
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)
library(scales)
library(changepoint)

source("../site_theme.R")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/carrier.duckdb", read_only = TRUE)

trend <- dbGetQuery(con, "
  SELECT
      period_start,
      MEDIAN(avg_d_mbps) AS median_dl,
      PERCENTILE_CONT(0.10) WITHIN GROUP (ORDER BY avg_d_mbps) AS p10_dl,
      PERCENTILE_CONT(0.25) WITHIN GROUP (ORDER BY avg_d_mbps) AS p25_dl,
      PERCENTILE_CONT(0.75) WITHIN GROUP (ORDER BY avg_d_mbps) AS p75_dl
  FROM ookla_clean
  GROUP BY period_start
  ORDER BY period_start
")
trend$period_start <- as.Date(trend$period_start)
```

## Change-Point Detection on Aggregate Throughput

`changepoint::cpt.mean()` tests whether the mean of a time series shifts at
any point. A detected change-point suggests a structural shift — which could
be a carrier policy change, a spectrum auction result, or network investment.

```{r}
#| label: changepoint-analysis
#| fig-cap: "Change-point detection on median US mobile download speed"
speed_ts <- trend$median_dl

cpt_result <- cpt.mean(speed_ts, method = "PELT", penalty = "BIC")
cpt_idx    <- cpts(cpt_result)

cat("Change points detected at index:", cpt_idx, "\n")
if (length(cpt_idx) > 0) {
  cat("Corresponding dates:", format(trend$period_start[cpt_idx]), "\n")
}

# Plot
trend$segment <- cumsum(c(0, diff(seq_along(speed_ts)) %in% cpt_idx))
ggplot(trend, aes(x = period_start, y = median_dl)) +
    geom_ribbon(aes(ymin = p25_dl, ymax = p75_dl), alpha = 0.15, fill = SITE_BLUE) +
    geom_line(linewidth = 0.8, color = SITE_BLUE) +
    geom_point(size = 2.5,    color = SITE_BLUE) +
    {if (length(cpt_idx) > 0)
        geom_vline(xintercept = trend$period_start[cpt_idx],
                   linetype = "dashed", color = "red", linewidth = 0.8)
    } +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Change-Point Detection: US Mobile Median Download Speed",
        subtitle = "Red dashed lines = detected structural breaks (PELT/BIC)",
        x = NULL, y = "Median Download (Mbps)"
    ) +
    theme_site()
```

## Variance Change Detection

Speed improvements can mask increasing *variance* — meaning some users are
getting much faster while others stay slow (consistent with throttling behavior).

```{r}
#| label: variance-changepoint
#| fig-cap: "Change-point detection on speed variance"
var_ts <- dbGetQuery(con, "
  SELECT period_start, STDDEV(avg_d_mbps) AS sd_dl
  FROM ookla_clean
  GROUP BY period_start
  ORDER BY period_start
")$sd_dl

cpt_var <- cpt.var(var_ts, method = "PELT", penalty = "BIC")
cat("Variance change points at indices:", cpts(cpt_var), "\n")
```

## Simulating Per-User Throttle Signatures

To understand what throttle detection looks like at the user level, we simulate
a throughput time series that crosses a throttle threshold:

```{r}
#| label: fig-throttle-sim
#| fig-cap: "Simulated user throughput with throttle at 50 GB threshold"
set.seed(42)
n_days     <- 30
threshold  <- 50   # GB
gb_per_day <- 2.5  # average daily consumption

usage <- cumsum(rnorm(n_days, mean = gb_per_day, sd = 0.4))
speed <- ifelse(usage < threshold,
                rnorm(n_days, mean = 80, sd = 8),   # pre-throttle: ~80 Mbps
                rnorm(n_days, mean = 5,  sd = 1))   # post-throttle: ~5 Mbps

sim <- data.frame(
    day   = 1:n_days,
    usage = usage,
    speed = pmax(speed, 0.5)
)
throttle_day <- which(usage >= threshold)[1]

ggplot(sim, aes(x = day, y = speed)) +
    geom_line(color = SITE_BLUE, linewidth = 0.8) +
    geom_point(size = 2, color = SITE_BLUE) +
    geom_vline(xintercept = throttle_day, linetype = "dashed",
               color = SITE_CORAL, linewidth = 0.9) +
    annotate("text", x = throttle_day + 0.5, y = max(sim$speed) * 0.9,
             label = paste0("Threshold crossed\n(day ", throttle_day, ")"),
             hjust = 0, size = 3.5, color = SITE_CORAL) +
    scale_y_continuous(labels = label_number(suffix = " Mbps")) +
    labs(
        title = "Simulated User Throughput with Data Cap Throttle",
        subtitle = paste0("50 GB threshold; ~", round(gb_per_day, 1), " GB/day usage"),
        x = "Day of Month", y = "Download Speed (Mbps)"
    ) +
    theme_site()
```

## Regression Discontinuity Design

If you know the stated threshold (e.g. 50 GB), RDD is the cleanest estimator
of the causal throttle effect — it compares users just below vs. just above
the limit:

```{r}
#| label: rdd-concept
# Conceptual RDD setup (requires per-user data from Wehe/M-Lab)
# With: df having columns `usage_gb` and `speed_mbps`
#
# library(rdrobust)
# rdd_result <- rdrobust(y = df$speed_mbps,
#                        x = df$usage_gb,
#                        c = 50)          # cutoff = 50 GB
# summary(rdd_result)
#
# The RDD estimator gives the discontinuous jump in speed at the threshold.
# A large negative jump is evidence of a hard throttle policy.

cat("RDD requires per-user data. See next steps for Wehe/M-Lab integration.")
```

## Next Steps: Wehe and M-Lab

The Ookla tile data establishes the market-level baseline. To detect
app-specific or user-level throttling:

| Dataset | How to access | What it adds |
|---|---|---|
| **Wehe** | `https://dd.meddle.mobi/wehe.html` | Per-test throttle flag, app-specific, carrier metadata |
| **M-Lab NDT** | BigQuery `measurement-lab.ndt.unified_downloads` | Per-test download speed with ASN (carrier) |
| **FCC MBA** | `data/fcc/` (from `fetch_fcc.R`) | Panel tests on fixed devices; known carrier |

```r
# M-Lab via BigQuery (requires Google Cloud free tier)
# library(bigrquery)
# con_bq <- dbConnect(bigquery(),
#                     project = "your-project-id",
#                     dataset = "measurement-lab.ndt")
```

## Problems

1. Apply `cpt.meanvar()` instead of `cpt.mean()` — does it detect additional
   breakpoints? What does a mean+variance change suggest vs. mean-only?

2. Look up the dates of the T-Mobile/Sprint merger completion and T-Mobile's
   5G mid-band rollout milestones. Do any detected change-points align?

3. Design a study using the Wehe dataset to test whether YouTube traffic is
   throttled differently than a generic HTTPS download on the same carrier.
   What statistical test would you use? What confounders would you need to
   control for?
